---
title: "WRFmet"
author: "Mazama Science"
date: "8/19/2020"
output: html_document
vignette: >
  %\VignetteIndexEntry{WRFmet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading WRF Data

Loading WRF data can be easily done using WRFmet's `wrf_load` function. This 
function accesses WRF model run files ...[in progress]...

[Explain WRF data dir]

```{r, load_example}
library(WRFmet)

# raster <- wrf_load(
#   modelName = ...,
#   modelRun = ...,
#   modelMode = ...,
#   vars = c('HGT', 'TSK'),
#   res = 0.1
# )
```

You can also load in existing WRF NetCDF files from anywhere on your machine by 
specifying the `localPath` argument in `wrf_load()`. For example:

```
raster <- wrf_load(
  localPath = '~/Data/WRF/wrfout_d3-2020071512-f07-0000.nc',
  vars = c('HGT', 'TSK'),
  res = 0.1
)
```

## Plotting WRF Data

WRFmet provides several preset plot types for visualizing raster, vector, and 
spatial point/polygon data. These include:

...

These plots are preset combinations of various building-block "layers". Custom
plots can be built by stacking these layers which function similar to GGPlot's
`geom_*` elements. This process is covered in further detail in the "Static 
Layered Plots" article.
