---
title: "AirFireWRF"
author: "Mazama Science"
date: "8/19/2020"
output: html_document
vignette: >
  %\VignetteIndexEntry{AirFireWRF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## WRF Model Run Files

The USFS [AirFire](https://www.airfire.org) team uses meteorological forecasts 
produced by the Weather Research & Forecasting (WRF) model. These "model runs" 
predict atmospheric conditions over the US for several dozen hours ahead. Output
from each WRF model run is stored as a collection of files containing data for 
individual forecast hours. For instance, a file named `PNW-4km_2020082712_07.nc` 
would describe the PNW-4km atmosphere forecasted 7 hours after the start of the 
2020-08-27 12pm model run. 

The **AirFireWRF** package allows users to download, read, and visualize the 
contents of these individual WRF data files.

_NOTE:  Because WRF output files are very large, R code in this vignette is not 
run when the package is built. This document contains output from commands run 
on August 31, 2020._

## Loading WRF Data

In order to start analyzing WRF output, we should first determine which model 
runs are available through AirFire. This can be done using the 
`wrf_latestModelRun()` function, in which we specify a model domain and the 
maximum number of recent model runs to be listed (defaults to 1):

```
library(AirFireWRF)

modelRun <- wrf_latestModelRun(modelName = "PNW-4km")
modelRun
```

```
[1] "2020083112"
```

The most recent model run was initialized at 
`"2020-08-31 12:00:00 UTC"`. Before we try to download files from that run 
though, we should first define a directory to hold all our WRF data. A file for 
a single forecast hour can be upwards of a hundred Megabytes, so a whole model 
run (AirFire's are typically 84 hours) could take up quite a bit of space. 
Therefore, it is nice to specify a dedicated place to hold all this information, 
and we can do that using AirFireWRF's `setWRFDataDir()` function:

```
setWRFDataDir("~/Data/WRF")
```

Now let's request one of the forecasted hours from the 
`"2020-08-31 12:00:00 UTC"` model run. Running `wrf_download()` will give us 
a file name which we can use for later, and if the file already exists on our 
machine then it will not be downloaded again.

```
wrfFile <- wrf_download(
  modelName = "PNW-4km",
  modelRun = modelRun,
  modelRunHour = 7       # 7 is the first hour available in most model runs
)

print(wrfFile)
```

```
[1] "/Users/jonathan/Data/WRF/PNW-4km_2020083112_07.nc"
```

We can see what WRF files we've already downloaded into our `WRFDataDir/` with
`wrf_downloaded()`:

```
wrf_downloaded()
```

```
[1] "PNW-4km_2020083112_07.nc"
```

So now we have access to a single hour in a WRF model run. This file holds 
readings for over a hundred different variables in a grid across the PNW-4km 
domain. We can load this data in as a raster object to start analyzing and 
plotting its contents. In addition to the model run identifiers (domain, run, 
hour), we must also specify what variables we'd like to include and -- if we 
choose  --  the resolution and extent of the raster that will be created.

```
raster <- wrf_load(
  modelName = "PNW-4km",
  modelRun = modelRun,
  modelRunHour = 7,
  varName = c("HGT", "TSK"), # Meter elevation (HGT) and Kelvin temperature (TSK)
  res = 0.1               # Grid cells will be .1x.1 degrees lon/lat
)
```

If you would like to simply provide the path of a WRF file, you can pass that 
in with `wrf_load`'s `localPath` argument. If the specified model run hour is 
not on your machine, it will be downloaded behind the scenes if it is available
from AirFire.

```
raster <- wrf_load(
  localPath = '~/some/other/dir/PNW-4km_2020071512_07.nc',
  varName = c("HGT", "TSK"),
  res = 0.1
)
```

The output of `wrf_load()` is a **raster** package _RasterBrick_ object with 
a separate _RasterLayer_ for each specified variable.

```
print(raster)
```

```
class      : RasterBrick 
dimensions : 111, 234, 25974, 2  (nrow, ncol, ncell, nlayers)
resolution : 0.1001554, 0.1000347  (x, y)
extent     : -130.4868, -107.0504, 39.49399, 50.59784  (xmin, xmax, ymin, ymax)
crs        : +proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0 
source     : memory
names      :           HGT,           TSK 
min values : -5.960464e-08,  2.731600e+02 
max values :     3571.1672,      317.7452 
```

## Plotting WRF Data

Let's find out what these atmospheric readings actually look like. AirFireWRF 
provides several preset plotting functions for visualizing this information as 
rasters and vector fields with spatial point/polygon overlays. These functions
include:

* `wrf_rasterPlot()` - Plots a single _RasterLayer_.
* `wrf_standardPlot()` - Plots a _RasterLayer_, _SpatialPolygonsDataFrame_, 
_SpatialPointsDataFrame_, and a vector field, in that order. Any of these can be 
missing, so this is more of a catch-all plot that can do the same as 
`wrf_rasterPlot()`, just with many more options.

So let's throw together a plot for the elevation (HGT) field of our raster:

_NOTE: In order to build this vignette, this plot will display an example raster
provided by AirFireWRF package "example_PNW"._

```{r}
library(AirFireWRF)
library(raster)

wrf_rasterPlot(
  raster = AirFireWRF::example_PNW$HGT,
  varName = "HGT",
  title = "PNW Elevation",
  flab = "Elev (m)",
  ratio = 1.4
)
```

`varName` specifies which *RasterLayer* in a *RasterBrick* to display. You can 
also simply provide a single *RasterLayer* like so:

```
wrf_rasterPlot(
  raster = AirFireWRF::example_PNW$HGT,
  title = "PNW Elevation",
  flab = "Elev (m)",
  ratio = 1.4
)
```

If we want to create a more complex plot, the `plot_standard()` function
provides a full-featured tool for visualizing WRF data and other spatial 
information. It builds a multi-layered plot including (in order, if the data is 
provided):

1. A background raster layer
2. A spatial polygons layer
3. A spatial points layer
4. A U.S. states layer
5. A vector field layer

Let's try building a plot of the Pacific Northwest with:

1. An elevation background raster
2. State borders outlined in red
3. A wind vector field

```{r}
wrf_standardPlot(
  raster = example_PNW,
  bgName = "HGT",
  uName = "U10",
  vName = "V10",
  states = TRUE,
  stateColor = "red",
  stateFill = "transparent",
  arrowColor = "black",
  title = "PNW Elevation & Wind Velocity",
  flab = "Elev (m)",
  xlim = c(-125, -111),
  ylim = c(42, 49),
  ratio = 1.4
)
```

These plots are actually preset combinations of various building-block "layers". 
Custom plots can be built by stacking these layers which function similar to 
**ggplot**'s `geom_*` elements. This process is covered in further detail in the 
"Static Layered Plots" article.
